---
title: "Problem set 3"
author: "Murphy Li"
date: "2023-04-16"
output: pdf_document
---

```{r}
library(tidyverse)
library(gridExtra)
library(lmtest)
library(stargazer)
library(sandwich)

df<-read.csv('E:/Data analysis practice/R/CBR data/P_Data_Extract_From_World_Development_Indicators/Data.csv')
```


Plotting each with CBR
```{r}
#remove na, find year with most observations after omiting na
df<-df%>%
  na.omit()

df%>%
  count(Year)


# 2015 has the least entries with na


df%>%
  filter(Year==2015)%>%
  ggplot(aes(x=UHC_index, y=Crude_birth_rate))+
  geom_point()+
  xlab('Universal Heathcare Index')+ ylab('Crude Birth Rate')+ labs(title='UHC Index v CBR in 2015')+
  theme_bw()+
  geom_smooth(se=F)+
  ggsave('UHC index v CBR.png', width=7, height=7/1.618)


df%>%
  filter(Year==2015)%>%
  ggplot(aes(x=GNI_percap, y=Crude_birth_rate))+
  geom_point()+
  xlab('GNI Per Capita (USD)')+ ylab('Crude Birth Rate')+ labs(title='GNI Per cap v CBR in 2015')+
  theme_bw()

#using log on GNI per cap
df<-df%>%
  mutate('log_GNI_percap'=log(GNI_percap))

df%>%
  filter(Year==2015)%>%
  ggplot(aes(x=log_GNI_percap, y=Crude_birth_rate))+
  geom_point()+
  xlab('Logged GNI Per Capita (USD)')+ ylab('Crude Birth Rate')+ labs(title='Logged GNI Per cap v CBR in 2015')+
  theme_bw()+
  geom_smooth(se=F)+
  ggsave('income v cbr.png', width=7, height=7/1.618)

df%>%
  filter(Year==2015)%>%
  ggplot(aes(x=Gov_edu_exp_pergdp, y=Crude_birth_rate))+
  geom_point()+
  xlab('Government Spending on Education (% GDP)')+ ylab('Crude Birth Rate')+   labs(title='Gov Education Spending v CBR in 2015')+
  theme_bw()+
  ggsave('education v cbr.png', width=7, height=7/1.618)

df%>%
  filter(Year==2015)%>%
  ggplot(aes(x=Fe_lab_force_part_rate_a, y=Crude_birth_rate))+
  geom_point()+
  xlab('Female Labor Force Participation rate (%)')+ ylab('Crude Birth Rate')+ labs(title='Female Labor Force Participation rate  v CBR in 2015')+
  theme_bw()+
  ggsave('fe labor force pr v cbr.png', width=7, height=7/1.618)

#summary of data 
stargazer(df)


#OLS model
model<-lm(Crude_birth_rate~UHC_index+log_GNI_percap+Gov_edu_exp_pergdp+Fe_lab_force_part_rate_a, df[df$Year==2015,])
summary(model)
stargazer(model, title='OLS Regression Results', covariate.labels=c("Universal Heatlh Care Index","Logged GNI Per Capita","Government Spending on Education (% GDP)", "Female Labor Force Participation Rate"),dep.var.labels=c("Crude Birth Rate"))

#ols less non significant varibale (Female laborforce)
model2<-lm(Crude_birth_rate~UHC_index+log_GNI_percap+Gov_edu_exp_pergdp, df[df$Year==2015,])
stargazer(model, model2, title='OLS Regression Results', covariate.labels=c("Universal Heatlh Care Index","Logged GNI Per Capita","Government Spending on Education (% GDP)"),dep.var.labels=c("Crude Birth Rate"))

```
It appears that UHC and GNI per cap has a negative relationship with CBR (as expected from theory), surprisingly, not much can be said about female labor force participation rate and government spending on education

The model I chose is a linear regression model with OLS. The mathematical formula of this prediction function is

$$ Crude Birth Rate= 59-0.47 X_1 - 1.57X_2 +0.67 X_3 + 0.04 X_4 $$
Where 
\begin{center}
$X_1=$ UHC Index\\
$X_2=$ logged GNI per Capita \\
$X_3=$ Government expenditure on education as a \% of GDP\\
$X_4=$ Female labor force participation rate\\
\end{center}

Breusch pagan test
```{r}
bptest(model)

bptest(model2)

```

robust standard errors
```{r}

coeftest(model, vcov = vcovHC(model, type = 'HC0'))
coeftest(model2, vcov = vcovHC(model2, type = 'HC0'))
```


Heteroskedasticity (see if variance is constant) graph 
```{r}
res <- data.frame(resid(model),fitted(model))

res%>%
  ggplot(aes(x=fitted.model., y=resid.model.))+
  geom_point()+
  theme_bw()+
  xlab('Fitted Values')+ylab('Residuals')+
  geom_hline(yintercept=0)+
  ggsave('Residual plot ols.png',width=7, height=7/1.618)

```

Weight least squares 
```{r}
#wls with all var
weight2 <- 1 / lm(abs(model$residuals) ~ model$fitted.values)$fitted.values^2
model_rr2<-lm(Crude_birth_rate~UHC_index+log_GNI_percap+Gov_edu_exp_pergdp+Fe_lab_force_part_rate_a, df[df$Year==2015,], weights = weight2)

summary(model_rr2)

#wls without insignidicant var
weight <- 1 / lm(abs(model2$residuals) ~ model2$fitted.values)$fitted.values^2
model_rr<-lm(Crude_birth_rate~UHC_index+log_GNI_percap+Gov_edu_exp_pergdp, df[df$Year==2015,], weights=weight)

summary(model_rr)

stargazer(model_rr2, model_rr, title='WLS Regression Results', align=T, covariate.labels=c("Universal Heatlh Care Index","Logged GNI Per Capita","Government Spending on Education (% GDP)", "Female Labor Force Participation Rate"))
```

Heteroskedasticity in new model
```{r}
res <- data.frame(resid(model_rr),fitted(model_rr))

res%>%
  ggplot(aes(x=fitted.model_rr., y=resid.model_rr.))+
  geom_point()+
  theme_bw()+
  xlab('Fitted Values')+ylab('Residuals')+
  geom_hline(yintercept=0)+
  ggsave('Residual plot wls.png',width=7, height=7/1.618)
```




Using predict on 2017 data, plotting with x=UHC index y= modeled crude birth rate
```{r}
newdat<-df%>%
  filter(Year==2017)%>%
  dplyr::select(UHC_index,log_GNI_percap,Gov_edu_exp_pergdp,Fe_lab_force_part_rate_a)

res <- predict(model_rr, newdata = newdat,se.fit=T)
data.frame(res) %>%
  mutate(lower = fit - 1.96 * se.fit,
         upper = fit + 1.96 * se.fit,
         x=newdat$UHC_index) %>%
  dplyr::select(-residual.scale,-se.fit) %>%
  ggplot(aes(x=x, y=fit, ymin=lower, ymax=upper)) +
  geom_ribbon(alpha=0.2) +
  theme_bw() + 
  geom_point()+
  xlab('UHC Index')+ylab('Modeled Crude Birthrate')+labs(title='Modeled CBR with 2017 data')

data.frame(res) %>%
  mutate(lower = fit - 1.96 * se.fit,
         upper = fit + 1.96 * se.fit,
         x=newdat$log_GNI_percap) %>%
  dplyr::select(-residual.scale,-se.fit) %>%
  ggplot(aes(x=x, y=fit, ymin=lower, ymax=upper)) +
  geom_ribbon(alpha=0.2) +
  theme_bw() + 
  geom_point()+
  xlab('logged GNI per cap')+ylab('Modeled Crude Birthrate')+labs(title='Modeled CBR with 2017 data')

data.frame(res) %>%
  mutate(lower = fit - 1.96 * se.fit,
         upper = fit + 1.96 * se.fit,
         x=newdat$Gov_edu_exp_pergdp) %>%
  dplyr::select(-residual.scale,-se.fit) %>%
  ggplot(aes(x=x, y=fit, ymin=lower, ymax=upper)) +
  geom_ribbon(alpha=0.2) +
  theme_bw() + 
  geom_point()+
  xlab('Government spending on education (% GDP)')+ylab('Modeled Crude Birthrate')+labs(title='Modeled CBR with 2017 data')

data.frame(res) %>%
  mutate(lower = fit - 1.96 * se.fit,
         upper = fit + 1.96 * se.fit,
         x=newdat$Fe_lab_force_part_rate_a) %>%
  dplyr::select(-residual.scale,-se.fit) %>%
  ggplot(aes(x=x, y=fit, ymin=lower, ymax=upper)) +
  geom_ribbon(alpha=0.2) +
  theme_bw() + 
  geom_point()+
  xlab('Female labor force participation rate')+ylab('Modeled Crude Birthrate')+labs(title='Modeled CBR with 2017 data')
```


Using Predict, holding all covariates as mean except universal health care index ranging from the min and max of data set

```{r}
#UHC Index
xhyp<-min(df$UHC_index):max(df$UHC_index)
newdat<-data.frame(
  UHC_index=xhyp,
  log_GNI_percap=mean(df[df$Year==2015,]$log_GNI_percap),
  Gov_edu_exp_pergdp=mean(df[df$Year==2015,]$Gov_edu_exp_pergdp)
)

res <- predict(model2, newdata = newdat, se.fit=T)

data.frame(res) %>%
  mutate(lower = fit - 1.96 * se.fit,
         upper = fit + 1.96 * se.fit,
         x=xhyp) %>%
  dplyr::select(-residual.scale,-se.fit) %>%
  ggplot(aes(x=x, y=fit, ymin=lower, ymax=upper)) +
  geom_ribbon(alpha=0.2) +
  theme_bw() + 
  geom_line()+
  xlab('Universal Health Care Index')+ylab('Modeled Crude Birthrate')+
  labs(title = 'UHC Index')
  ggsave('predict hold const UHC index.png', width=7, height=7/1.618)

#logged GNI per Cap
xhyp<-min(df$log_GNI_percap):max(df$log_GNI_percap)
newdat<-data.frame(
  UHC_index=mean(df[df$Year==2015,]$UHC_index),
  log_GNI_percap=xhyp,
  Gov_edu_exp_pergdp=mean(df[df$Year==2015,]$Gov_edu_exp_pergdp)
)

res <- predict(model2, newdata = newdat, se.fit=T)

data.frame(res) %>%
  mutate(lower = fit - 1.96 * se.fit,
         upper = fit + 1.96 * se.fit,
         x=xhyp) %>%
  dplyr::select(-residual.scale,-se.fit) %>%
  ggplot(aes(x=x, y=fit, ymin=lower, ymax=upper)) +
  geom_ribbon(alpha=0.2) +
  theme_bw() + 
  geom_line()+
  xlab('Logged GNI Per Capita')+ylab('Modeled Crude Birthrate')+
  labs(title = 'Logged GNI Per Capita')+
  ggsave('predict hold const logged GNI.png', width=7, height=7/1.618)

#GNI per Cap
xhyp1<-min(df$GNI_percap):max(df$GNI_percap)
xhyp<-log(xhyp1)
newdat<-data.frame(
  UHC_index=mean(df[df$Year==2015,]$UHC_index),
  log_GNI_percap=xhyp,
  Gov_edu_exp_pergdp=mean(df[df$Year==2015,]$Gov_edu_exp_pergdp)
)

res <- predict(model2, newdata = newdat, se.fit=T)

data.frame(res) %>%
  mutate(lower = fit - 1.96 * se.fit,
         upper = fit + 1.96 * se.fit,
         x=xhyp1) %>%
  dplyr::select(-residual.scale,-se.fit) %>%
  ggplot(aes(x=x/1000, y=fit, ymin=lower, ymax=upper)) +
  geom_ribbon(alpha=0.2) +
  theme_bw() + 
  geom_line()+
  xlab('GNI Per Capita (USD in 1000s)')+ylab('Modeled Crude Birthrate')+
  labs(title = 'GNI Per Capita')+
  ggsave('predict hold const GNI.png', width=7, height=7/1.618)


#gov spending on edu
xhyp<-min(df$Gov_edu_exp_pergdp):max(df$Gov_edu_exp_pergdp)
newdat<-data.frame(
  UHC_index=mean(df[df$Year==2015,]$UHC_index),
  log_GNI_percap=mean(df[df$Year==2015,]$log_GNI_percap),
  Gov_edu_exp_pergdp=xhyp
)

res <- predict(model2, newdata = newdat, se.fit=T)

data.frame(res) %>%
  mutate(lower = fit - 1.96 * se.fit,
         upper = fit + 1.96 * se.fit,
         x=xhyp) %>%
  dplyr::select(-residual.scale,-se.fit) %>%
  ggplot(aes(x=x, y=fit, ymin=lower, ymax=upper)) +
  geom_ribbon(alpha=0.2) +
  theme_bw() + 
  geom_line()+
  xlab('Government Spending on Education (% GDP)')+ylab('Modeled Crude Birthrate')+
  labs(title = 'Government Spending on Education (% GDP)')+
  ggsave('predict hold const gov ed spending.png', width=7, height=7/1.618)



```

Bootstrapping UHC Index
```{r}
#first difference UHC index
diff <- rep(NA, 10000)
for(i in 1:10000){
  bootdat <- sample_n(df[df$Year==2015,],size=nrow(df[df$Year==2015,]), replace=T)
  reg.out <- lm(Crude_birth_rate~UHC_index+log_GNI_percap+Gov_edu_exp_pergdp, bootdat)
  newdat = data.frame(
    UHC_index = c(min(df$UHC_index[df$Year==2015]),max(df$UHC_index[df$Year==2015])),
    log_GNI_percap = mean(bootdat$log_GNI_percap),
    Gov_edu_exp_pergdp = mean(bootdat$Gov_edu_exp_pergdp)
  )
  yhat <- predict(reg.out,
                  newdata = newdat)
  diff[i] <- yhat[2]-yhat[1]
}

ggplot(as.data.frame(diff), aes(x=diff)) + geom_histogram(color='black', fill='white')+
  xlab('Difference')+ylab('Frequency')+labs(title='UHC Index First difference with 2015 data')+
  theme_bw()+
  geom_vline(aes(xintercept=mean(diff)),
            color="blue", size=0.75, linetype="dashed")+
  ggsave('UHC first diff.png', width=7, height=7/1.618)


bot.sum<-data.frame(
  Average=mean(diff),
  'Lower Quartile'=quantile(diff, 0.025),
  'Upper Quartile'=quantile(diff, 0.975)
)
stargazer(bot.sum, summary=FALSE, rownames=FALSE)




```

Bootstrapping logged gni per cap
```{r}
#first difference logged gni per cap
diff <- rep(NA, 10000)
for(i in 1:10000){
  bootdat <- sample_n(df[df$Year==2015,],size=nrow(df[df$Year==2015,]), replace=T)
  reg.out <- lm(Crude_birth_rate~UHC_index+log_GNI_percap+Gov_edu_exp_pergdp, bootdat)
  newdat = data.frame(
    UHC_index = mean(bootdat$UHC_index),
    log_GNI_percap = c(min(df$log_GNI_percap[df$Year==2015]),max(df$log_GNI_percap[df$Year==2015])),
    Gov_edu_exp_pergdp = mean(bootdat$Gov_edu_exp_pergdp)
  )
  yhat <- predict(reg.out,
                  newdata = newdat)
  diff[i] <- yhat[2]-yhat[1]
}

ggplot(as.data.frame(diff), aes(x=diff)) + geom_histogram(color='black', fill='white')+
  xlab('Difference')+ylab('Frequency')+labs(title='Logged GNI Per Capita First difference with 2015 data')+
  theme_bw()+
  geom_vline(aes(xintercept=mean(diff)),
            color="blue", size=0.75, linetype="dashed")+
  ggsave('GNI per cap first diff.png', width=7, height=7/1.618)


bot.sum<-data.frame(
  Average=mean(diff),
  'Lower Quartile'=quantile(diff, 0.025),
  'Upper Quartile'=quantile(diff, 0.975)
)
stargazer(bot.sum, summary=FALSE, rownames=FALSE)



```

Bootstrapping gov spending on education
```{r}
#first difference gov spending on education
diff <- rep(NA, 10000)
for(i in 1:10000){
  bootdat <- sample_n(df[df$Year==2015,],size=nrow(df[df$Year==2015,]), replace=T)
  reg.out <- lm(Crude_birth_rate~UHC_index+log_GNI_percap+Gov_edu_exp_pergdp, bootdat)
  newdat = data.frame(
    UHC_index = mean(bootdat$UHC_index),
    log_GNI_percap = mean(bootdat$log_GNI_percap),
    Gov_edu_exp_pergdp =c(min(df$Gov_edu_exp_pergdp[df$Year==2015]),max(df$Gov_edu_exp_pergdp[df$Year==2015]))
  )
  yhat <- predict(reg.out,
                  newdata = newdat)
  diff[i] <- yhat[2]-yhat[1]
}

ggplot(as.data.frame(diff), aes(x=diff)) + geom_histogram(color='black', fill='white')+
  xlab('Difference')+ylab('Frequency')+labs(title='Government Spending on Education First difference with 2015 data')+
  theme_bw()+
  geom_vline(aes(xintercept=mean(diff)),
            color="blue", size=0.75, linetype="dashed")+
  ggsave('Gov edu spending first diff.png', width=7, height=7/1.618)


bot.sum<-data.frame(
  Average=mean(diff),
  'Lower Quartile'=quantile(diff, 0.025),
  'Upper Quartile'=quantile(diff, 0.975)
)
stargazer(bot.sum, summary=FALSE, rownames=FALSE)



```
Bootstrapping female lab part rate

```{r}
#first difference female lab part rate
diff <- rep(NA, 10000)
for(i in 1:10000){
  bootdat <- sample_n(df[df$Year==2015,],size=nrow(df[df$Year==2015,]), replace=T)
  reg.out <- rlm(Crude_birth_rate~UHC_index+log_GNI_percap+Gov_edu_exp_pergdp+Fe_lab_force_part_rate_a, bootdat)
  newdat = data.frame(
    UHC_index = mean(bootdat$UHC_index),
    log_GNI_percap = mean(bootdat$log_GNI_percap),
    Gov_edu_exp_pergdp =mean(bootdat$Gov_edu_exp_pergdp),
    Fe_lab_force_part_rate_a=c(min(df$Fe_lab_force_part_rate_a[df$Year==2015]),max(df$Fe_lab_force_part_rate_a[df$Year==2015]))
  )
  yhat <- predict(reg.out,
                  newdata = newdat)
  diff[i] <- yhat[2]-yhat[1]
}

ggplot(as.data.frame(diff), aes(x=diff)) + geom_histogram(color='black', fill='white')+
  xlab('Difference')+ylab('Frequency')+labs(title='Feamle Labor Force Pariticipation rate First difference with 2015 data')+
  theme_bw()+
  geom_vline(aes(xintercept=mean(diff)),
            color="blue", size=0.75)+
  ggsave('female labor part rate first diff.png', width=7, height=7/1.618)


data.frame(
  Average=mean(diff),
  lower_quartile=quantile(diff, 0.025),
  upper_quartile=quantile(diff, 0.975)
)



```


using same year as model data 2015
```{r}
diff <- rep(NA, 10000)
for(i in 1:10000){
  bootdat <- sample_n(df[df$Year==2015,],size=nrow(df[df$Year==2015,]), replace=T)
  reg.out <- lm(Crude_birth_rate~UHC_index+log_GNI_percap+Gov_edu_exp_pergdp+Fe_lab_force_part_rate_a, bootdat)
  newdat = data.frame(
    UHC_index = mean(bootdat$UHC_index),
    log_GNI_percap = mean(bootdat$log_GNI_percap),
    Gov_edu_exp_pergdp = mean(bootdat$Gov_edu_exp_pergdp),
    Fe_lab_force_part_rate_a = mean(bootdat$Fe_lab_force_part_rate_a)
  )
  yhat[1] <- predict(reg.out,
                  newdata = newdat)
  yhat[2] <- predict(model,
                  newdata = newdat)
  diff[i] <- yhat[2]-yhat[1]
}

ggplot(as.data.frame(diff), aes(x=diff)) + geom_histogram(color='black', fill='white')+
  xlab('Difference in prediction')+ylab('Frequency')+labs(title='Difference in predicted values for 2015')

data.frame(
  y=mean(diff),
  l=quantile(diff, 0.025),
  u=quantile(diff, 0.975)
)

```

Using 2017 (which is not what the model was based on)
```{r}
diff <- rep(NA, 10000)
for(i in 1:10000){
  bootdat <- sample_n(df[df$Year==2017,],size=nrow(df[df$Year==2017,]), replace=T)
  reg.out <- lm(Crude_birth_rate~UHC_index+log_GNI_percap+Gov_edu_exp_pergdp+Fe_lab_force_part_rate_a, bootdat)
  newdat = data.frame(
    UHC_index = mean(bootdat$UHC_index),
    log_GNI_percap = mean(bootdat$log_GNI_percap),
    Gov_edu_exp_pergdp = mean(bootdat$Gov_edu_exp_pergdp),
    Fe_lab_force_part_rate_a = mean(bootdat$Fe_lab_force_part_rate_a)
  )
  yhat[1] <- predict(reg.out,
                  newdata = newdat)
  yhat[2] <- predict(model,
                  newdata = newdat)
  diff[i] <- yhat[2]-yhat[1]
}

ggplot(as.data.frame(diff), aes(x=diff)) + geom_histogram(color='black', fill='white')+
  xlab('Difference in prediction')+ylab('Frequency')+labs(title='Difference in predicted values for 2017')
data.frame(
  y=mean(diff),
  l=quantile(diff, 0.025),
  u=quantile(diff, 0.975)
)
```
Using data from all years (which is not what the model was based on)
```{r}
diff <- rep(NA, 10000)
for(i in 1:10000){
  bootdat <- sample_n(df,size=nrow(df), replace=T)
  reg.out <- lm(Crude_birth_rate~UHC_index+log_GNI_percap+Gov_edu_exp_pergdp+Fe_lab_force_part_rate_a, bootdat)
  newdat = data.frame(
    UHC_index = mean(bootdat$UHC_index),
    log_GNI_percap = mean(bootdat$log_GNI_percap),
    Gov_edu_exp_pergdp = mean(bootdat$Gov_edu_exp_pergdp),
    Fe_lab_force_part_rate_a = mean(bootdat$Fe_lab_force_part_rate_a)
  )
  yhat[1] <- predict(reg.out,
                  newdata = newdat)
  yhat[2] <- predict(model,
                  newdata = newdat)
  diff[i] <- yhat[2]-yhat[1]
}

ggplot(as.data.frame(diff), aes(x=diff)) + geom_histogram(color='black', fill='white')+
  xlab('Difference in prediction')+ylab('Frequency')+labs(title='Difference in predicted values from years 2013, 2015, 2017')

data.frame(
  y=mean(diff),
  l=quantile(diff, 0.025),
  u=quantile(diff, 0.975)
)
```
From bootstrapping, the model appears to be consistent and accurate


Using predict to predict CBR using data from all years, and look at difference (predicted-actual)
```{r}
#2015
df2015<-df%>%
  filter(Year==2015)
predicted2015<-predict(model, df2015)

error2015<-df2015$Crude_birth_rate-predicted2015
ggplot(as.data.frame(error2015), aes(x=error2015)) + geom_histogram(color='black', fill='white')+
xlab('Error term')+ylab('Frequency')+labs(title='Distribution of error terms in 2015')

data.frame(Mean=mean(error), Standard_deviation=sd(error), Variance=var(error))

#2017
df2017<-df%>%
  filter(Year==2017)
predicted2017<-predict(model, df2017)

error<-df2017$Crude_birth_rate-predicted2017
ggplot(as.data.frame(error), aes(x=error)) + geom_histogram(color='black', fill='white')+
xlab('Error term')+ylab('Frequency')+labs(title='Distribution of error terms in 2017')

data.frame(Mean=mean(error), Standard_deviation=sd(error), Variance=var(error))

#2019
df2019<-df%>%
  filter(Year==2019)
predicted2019<-predict(model, df2019)

error<-df2019$Crude_birth_rate-predicted2019
ggplot(as.data.frame(error), aes(x=error)) + geom_histogram(color='black', fill='white')+
xlab('Error term')+ylab('Frequency')+labs(title='Distribution of error terms in 2019')

data.frame(Mean=mean(error), Standard_deviation=sd(error), Variance=var(error))
```
The distribution of error terms look similar for all years, could indicate consistency in the model









